\section{Introduction}

The NSF-DOE Vera C.\ Rubin Observatory is designed to take data for the Legacy Survey of Space and Time over a period of 10 years.\cite{2019ApJ...873..111I}.
We made our first data preview release (DP1) in 2025 using data from our commissioning camera\cite{RTN-095} and we are currently preparing our second data preview (DP2) using data from LSSTCam.\cite{10.71929/rubin/2571927,RTN-111}
For DP1 our image data products were written in FITS format following the FITS standard\cite{FITS4} but the FITS standard is not in itself sufficient to fully define all the information that we need to record in our FITS headers.
The resultant FITS files were able to represent concepts such as variances and masks but provenance and point-spread function definitions were represented as opaque data that required specific Rubin science pipeline software\cite{PSTN-019} to interpret.
In this paper we will describe the FITS layout we used for DP1 and describe its limitations.
Then we will present our new approach to modeling our data structures and how we represent it in both FITS and HDF5 files.

\section{The Rubin Visit Image Data Model}

The most complex image dataset type we write are those from our visit images.
At Rubin we define ``visit'' as an on-sky exposure.
During construction we had planned to support an observing mode where two on-sky exposures would be taken sequentially without moving the telescope or instrument rotator and those would be called a visit.
This changed during commissioning when we decided to drop the two-exposure mode and so now a visit is always a single on-sky exposure.
For a visit observation the resulting files have to describe the detector properties, the point-spread function, the non-linear large-scale WCS, aperture corrections and other calibration effects.
Every visit includes a 32-bit mask as well as a variance array to go with the image data and FITS-style (keyword, value, and comment) metadata.

The data model consists of the following additional components.

\subsection{Bounded Fields}

Some of the data model components rely on being able to represent a mathematical function to be evaluated over specific regions of the image.
we call these \emph{bounded fields} and they are usually represented as Chebyshev polynomials but can also be specified as splines or lazy arithmetic combinations (e.g. sums or products) of multiple other bounded fields.

\subsection{Visit Information}

A \texttt{VisitInfo} contains standardized information about the entire visit.
This includes the weather conditions, the instrument and observatory location, the time of the observation, the boresight tracking coordinates, and the exposure time.
Confusingly, calibration observations do result in a \texttt{VisitInfo} being defined although some of the components are undefined in that case.
We will not consider calibration datasets any further in this discussion since they generally are representable by a subset of the visit data model.

\subsection{Filter Label}

This is a small look up table defining the physical filter used for the observation and the waveband for that filter.

\subsection{Detector Information}

LSSTCam has 189 science detectors along with additional wave front sensors and guider detectors.
Each dataset contains a special detector object that describes the properties of this detector such as the serial number and manufactuer, the amplifier layout, and how the cordinates transform from focal plane position in millimeters to pixel coordinate.

\subsection{Point-Spread Function}

There are many ways to represent a point-spread function (PSF) and we use a plugin system to allow different variants to be stored.
We use both a heavily modified version of \texttt{PSFEx} \cite{2011ASPC..442..435B,2013ascl.soft01001B} and \texttt{Piff} \cite{2021ascl.soft02024J,2021MNRAS.501.1282J} in our production pipelines.
The PSF can vary across the image and the PSF model has to be able to account for that.

\subsection{Exposure Summary Statistics}

During the analysis of a visit the pipeline calculates derived scalar quantities which we store in a small class.
These include summaries of global PSF properties, source counts, and effective exposure times and bounding boxes.
This information can be used by downstream pipeline components to decide whether a visit should be included in a co-add or not.
Unlike \texttt{VisitInfo} the summary statistics can change depending on pipeline configuration.

\subsection{World Coordinate Systems}

At the large scale of the LSSTCam focal plane combined with atmospheric effects the standard FITS approaches to specifying a world coordinate system are not sufficiently accurate.
We use the Starlink AST WCS library \cite{2016A&C....15...33B} as this gives us the flexibility to define our WCS in terms of individual steps that can be chained together in whatever order we deem appropriate.
In that way we can include large scale polynomial distortions as well as more localized effects by using splines.
The WCS is represented by an AST mapping that is abstracted from the user by a Rubin-specific interface.

\subsection{Validity Region}

This is a polygon in pixel coordinates defining the valid region of the image.
For most detectors, the valid polygon is just the detector's bounding box, but those on the edge of the focal plane have a valid polygon boundary that marks where vignetting occurs.

\subsection{Aperture Corrections}

Multiple aperture corrections can be stored, each one being labeled and associated with a bounded field.
Different measurement algorithms can result in different aperture corrections and the user can select the appropriate one.

\subsection{Transmission Curve}

This component represents the throughput as a function of wavelength, generally using units of Angstroms.

\subsection{Photometric Calibration}

Part of the visit processing determines how the image should be corrected to convert detector units to nJy.
The calibration can vary across the field and is represented by a bounded field.

\subsection{Class Hierarchy}

In the LSST Science Pipelines every image dataset is represented by the \texttt{lsst.afw.image.Exposure} class in both C++ and Python.
An \texttt{Exposure} consists of a \texttt{MaskedImage} and an \texttt{ExposureInfo}.
The \texttt{MaskedImage} contains two \texttt{Image} objects (including the variance) and a \texttt{Mask} object, and bounding boxes and metadata.
One design decision that has caused confusion over the years is that you cannot associate world coordinates with a \texttt{MaskedImage} or \texttt{Image}, even though, conceptually, everyone expects such an entity to have a well-defined WCS.
Instead the WCS is associated with the \texttt{ExposureInfo} where the additional components described above are defined.

\section{DP1 FITS Files}

\begin{figure}[ht]
\begin{center}
\begin{small}
\begin{verbatim}
No.    Name      Ver    Type      Cards   Dimensions   Format
  0  PRIMARY       1 PrimaryHDU     722   ()
  1  IMAGE         1 CompImageHDU     72   (256, 250)   float32
  2  MASK          1 CompImageHDU     88   (256, 250)   int32
  3  VARIANCE      1 CompImageHDU     72   (256, 250)   float32
  4  ARCHIVE_INDEX    1 BinTableHDU     41   429R x 7C   [1J, 1J, 1J, 1J, 1J, 64A, 64A]
  5  FilterLabel    1 BinTableHDU     28   1R x 3C   [2X, 32A, 32A]
  6  Detector      1 BinTableHDU    115   1R x 22C   [1QA(7), 1J, 1J, 1QA(14), 1J, ..., 1J]
  7  TransformMap    1 BinTableHDU     33   411R x 5C   [1QA(10), 1QA(7), 1QA(10), 1QA(7), 1J]
  8  ExposureSummaryStats    1 BinTableHDU     18   412R x 1C   [1QB(3410)]
  9  Detector      2 BinTableHDU    201   16R x 38C   [3X, 1QA(3), 1J, ..., 1D, 1QA(2)]
 10  Polygon       1 BinTableHDU     21   5R x 2C   [1D, 1D]
 11  SkyWcs        1 BinTableHDU     21   1R x 2C   [1QB(16161), 1QB(10442)]
 12  PsfexPsf      1 BinTableHDU     52   1R x 9C   [1J, 1J, 1J, 1J, 1J, 1J, 1D, 1D, 1E]
 13  PsfexPsf      2 BinTableHDU     46   1R x 8C   [2J, 1J, 6D, 6D, 3J, 41334E, 2D, 2D]
 14  ApCorrMap     1 BinTableHDU     21   6R x 2C   [64A, 1J]
 15  ChebyshevBoundedField    1 BinTableHDU     41   5R x 6C   [1J, 1J, 1J, 1J, 1J, 1D]
 16  ChebyshevBoundedField    2 BinTableHDU     42   2R x 6C   [1J, 1J, 1J, 1J, 1J, 9D]
 17  PhotoCalib    1 BinTableHDU     36   1R x 5C   [1X, 1D, 1D, 1J, 1J]
\end{verbatim}
\end{small}
\end{center}
\label{fig:old-layout}
\caption{The structure of a slice of a representative FITS file using the legacy schema.}
\end{figure}


A summary of the layout of a FITS file produced for DP1 is shown in Fig.~\ref{fig:old-layout} and shows the extensive use of FITS binary tables.
The pixel data use standard FITS data arrays with compression to represent the primary image data, the variance, and the mask.
The mask image is a 2-d 32-bit integer array supporting 32 individual mask planes.
\emph{How do we know the names of each mask plane? Is it hardcoded in afw?}

The FITS primary header contains the standard header values from the raw data file\cite{CTN-004} but a visit image file also adds headers itself on serialization.
In general it is not possible to accurately represent the world coordinates of a visit image using the standard FITS-WCS approach.
We therefore store an approximation of the WCS in the headers in SIP form\cite{2005ASPC..347..491S} so that tools like Astropy\cite{2013A&A...558A..33A} and Firefly\cite{2019ASPC..521...32W} can see some coordinates.
For DP1 these approximations were good to a few milliarcseconds, i.e. around the astrometric precision of the true WCS, but this may not always be true in the future.

The \texttt{VisitInfo} struct also serializes into the FITS primary header.
We now consider this to be an anti-pattern because the keywords are not namespaced.
This can lead to confusion where a header from an instrument using \texttt{PRESSURE} to represent the air pressure in one unit will get a new value in a potentially different unit.
Even more dangerously, we write the midpoint of the observation using the standard \texttt{DATE-AVG} header and we force \texttt{TIMESYS} to TAI without attempting to adjust any other standard \texttt{DATE-} headers that might be in the file from the instrument raw headers using UTC.
This can lead to errors from any tooling that is trying to interpret headers using the standard.
If we were approaching this today we would store the \texttt{VisitInfo} in the header using an approach like \texttt{HIERARCH LSST VISITINFO DATE-AVG} and make it absolutely clear which headers were coming from the serialization and which were originally from the raw data.
We also write the detector name and detector serial ID into headers \texttt{DETNAME} and \texttt{DETSER}.

The remaining content is stored in FITS binary tables.

\subsection{Table Archive System}

The fundamental serialization approach for the DP1 data involves a generic persistence infrastructure that converts exposure components into rows of tables that can be stored in FITS.

Each component registers itself as a storable that implements the creation of a tabular version of the object that can be stored.
During serialization each component is allocated an integer archive ID that is referenced by a table of contents stored in the \texttt{ARCHIVE\_INDEX} extension.
The ID itself is stored in the primary FITS header with key \texttt{ARCHIVE\_INDEX\_<COMPONENT>}.

Each component is then represented in the \texttt{ARCHIVE\_INDEX} table indexed by the ID which refers to a later HDU and rows within that HDU.
If components share the same table schema they are written to the same HDU, which leads to a storage efficiency but does mean that each HDU can not be interpreted without the primary index.

\subsubsection{Component Serializations}

The components themselves are serialized in different ways.

\begin{description}

  \item[\textbf{FilterLabel}]

  This is a very simple table with two rows.
  One row indicates the wave band (and whether it is defined or not) and the other row contains the physical filter name (and its definedness state).

  \item[\textbf{Detector}]

  This is written as two catalogs.
  The first summarizes detector properties and the second describes amplifier properties with one row per amplifier.

  \item[\textbf{TransformMap}]

  This contains 2-D mappings define detector transformations from different frames.
  The mappings are byte string native serializations of the AST library mappings.

\item[\textbf{Polygon}]

Polygons are represented as a two column table containing the \texttt{x} and \texttt{y} vertices.

\item[\textbf{SkyWcs}]

This represents the pixel to sky transformations represented as a byte string native serialization of the AST library.
Optionally we can also store the FITS WCS approximation of the full mapping.

\item[\textbf{PhotoCalib}]

This table contains the scalar calibration information and references to archive IDs of bounded fields.

\item[\textbf{ApCorrMap}]

This is a simple table that associates the name of the correction with a reference to the bounded field.

\item[\textbf{ChebyshevBoundedField}]

These define the bounded fields referenced from other HDUs.
Each row defines the Chebyshev order, the bounding box for which the polynomial is valid and the polynomial coefficients.

\emph{Why do we end up with two HDUs with Chebyshev fields? Is it because there are different coefficient counts?}

\item[\textbf{ExposureSummaryStats}]

This is a Python data class.
All Python packages must serialize as strings which are stored in a single bytes column.
For this class the serialization uses YAML, but since this schema is also used by other components the HDU will likely contain serializations from other classes and the HDU name is going to depend on the order of serialization.

\item[\textbf{PSFs}]

The PSF component is polymorphic.
AFW only sees it through the persistable interface.
The actual FITS representation depends on the concrete PSF class and its registered factory.
For example, \texttt{PsfexPsf} from \texttt{meas\_extensions\_psfex} writes two catalogs and other PSF subclasses use their own schemas.

\end{description}

\section{Limitations of the DP1 Serialization}

We have been battling the limitations of this serialization framework for some time.\cite{DMTN-120,DMTN-251}
A key limitation for implementers was the difficulty in adding new components to the archive system.
The core of the system requires C++ and it was only relatively recently that we were able to include Python classes into the serialization mechanism.
Even then, each class is required to inherit from the \texttt{Storable} infrastructure in order to hook into the serialization system.
This adds an unnecessary burden on each class.
Furthermore, since we expect every image dataset we write to be using the AFW \texttt{Exposure} classes this means that whenever we need to add support for a new component, every dataset type sees that component even if it is not useful for that variant.

Whilst the implementation complexity is only of concern to developers and is a one-off cost, a more fundamental issue relates to the long-term usability of these files.
As can be seen from Fig.~\ref{fig:old-layout} the extension names do not necessarily tell you what the extension is associated with; for example, bounded fields are serialized as extensions but you do not know which component the bounded field is associated with.
As described previously, multiple components can also be written to the same HDU, with the \texttt{EXTNAME} referring to only one of the components out of convenience.
In order to understand what each extension is part of you must first read the index extension.

The complex logic effectively requires an end-user uses the \texttt{lsst.afw} Python classes to read the extensions.
Re-implementing the logic in Astropy is potentially possible with modern large-language model coding tools using the AFW code as a reference, but it would be preferable to avoid this situation.
Additionally, we are formally required to fully document how to read the files we ship to the community, and so are motivated to ship a file format that can be described relatively straightforwardly.

\emph{Didn't we have one component that was a pickle inside the binary table?}

The requirement to read the archive index does have other ramifications.
The system was designed in an era when there was an assumption that the file would always be available on a POSIX file system.
For DP1 we store the data in Google Cloud Storage and DP2 will be stored at SLAC but accessed remotely from Google\cite{2024SPIE13101E..2BO} over https.
When a user requests a component, such as the valid polygon, we would ideally like to retrieve only the HDU containing that information.
The simplest option would be to download the entire file to local storage and then read the component.
To avoid downloading megabytes of pixel data, instead what we have to do is construct fake single-pixel image HDUs for the image/variance/mask arrays and then download all the extension HDUs so we can write a file locally and use the AFW code to read the index and then construct the extension.
This is not efficient but is an inevitable outcome of the approach.

\section{Changing Approach for DP2}

Given the discussion above, we decided that for DP2 we would change the way we represent image datasets in Python and also change how we serialize them.
Our main requirements were:

\begin{itemize}

  \item The main implementation language should be Python rather than C++.
  \item Each image dataset type (such as visit images, difference images, co-adds) should be represented by its own Python class.
  This allows each dataset type to define only the components that are directly relevant to that type.
  It also allows more type safety in the pipelines rather than always assuming a generic image container.
  \item Informed by the experience with the Advanced Scientific Data Format (ASDF)\cite{2015A&C....12..240G}, the data models written to the file must be associated with public formal schemas.
  \item It should be possible to locate and read a component efficiently.
  \item People should be able to install software to read the files with our data models with only a \texttt{pip install} and the FITS files must be usable from Astropy and general viewing tools.
\end{itemize}

One additional input into the process was to make a conscious effort to distinguish the data model from the serialization format.
Separating the model from the file format has been a long-running debate in the community\cite{2019ASPC..521..733C} as we discuss the limitations of the FITS format\cite{2015A&C....12..133T} and the possibilities for being more flexible in terms of archival artifacts being produced.

\subsection{The New Data Model}

The most fundamental part of the data model is the \texttt{Image} class.
This contains a two-dimensional Numpy data array with a pixel origin and units, a projection that can map pixels to world coordinates, a generic metadata data class that can describe an observation (\emph{Maybe it was a mistake to add ObservationInfo to Image given the discussion above}) as well as some simple keyword/value metadata support.

There is also a \texttt{Mask} class which is similar to \texttt{Image} except it has an additional schema that describes the mask planes (bit numbers, mask name, mask description) and internally is represented as a three-dimensional array even though the external interface is a two-dimensional array.
The third dimension uses unsigned bytes to allow arbitrary numbers of mask planes to be stored efficiently in groups of eight without being constrained to jump from 32 mask planes directly to 64.

Additional components are represented by using Python Pydantic models\cite{2025zndo..15662245C} which provide validation and can define formal schemas.
Every part of the model uses Pydantic and we intend to generate JSON Schema to allow formal validation by third parties.
We have decided that where a model has an equivalent definition in the ASDF schema we will use a model that is compatible.
We do not directly use the ASDF schema definitions but we do define ASDF-compatible variants for dealing with Astropy quantities, units, N-dimensional arrays, time, and table columns.

We have reimplemented the AFW component classes in Python, including bounded fields, aperture correction maps, and PSFs and these are used for the new \texttt{VisitImage} class.
Additionally there is a \texttt{ColorImage} class that consists of three images (red, green, and blue) and a \texttt{CellCoadd} that represents a modern implementation of our cell co-add types.\cite{DMTN-294,SITCOMTN-137}
The system has been designed to allow flexibility in image classes but with a standardized serialization system.

\subsection{Serialization System}

A conscious design decision was to abstract the file output from the serialization infrastructure.

Every saveable Python type in the package (such as \texttt{Image}, \texttt{MaskedImage}, \texttt{VisitImage}, the various PSFs, transforms) has a paired serialization model — a Pydantic \texttt{ArchiveTree} subclass that mirrors the type's on-disk shape (e.g., \texttt{VisitImageSerializationModel} or \texttt{ColorImageSerializationModel}).
The Python serialization model class is what is then persisted.
The two are linked by a single class method defined on the Python class that returns the archive tree.
The argument to this method is a small Pydantic model that names ``out-of-band data'' (large arrays, tables, frame sets) for a particular backend, with one existing for each backend (JSON, FITS, and NDF), referring to locations in the JSON model or an HDU number in a FITS file, or the hierarchical path of an HDF5 file.

When a Python type is written, an \texttt{OutputArchive} for the backend is created and passed to that object's serialization method.
A new archive tree Pydantic model is created where scalar fields are stored directly but large numerical arrays and tables are extracted and then referenced from the model.
If an entity appears more than once in the model (such as a PSF) it is stored once and then referenced.
Once the model is transformed to JSON the binary side channel information is then stored in a backend-specific way.

On read the JSON is located using the backend-specific code and validated.
The model is then handed off to a deserialization method which resolves all the references to other parts of the file and reads those to reconstruct the Python type.

This design is what makes the three backends interchangeable.
A Python class doesn't know whether it lives in a FITS HDU list or an HDS structure; it only knows how to translate itself to and from a Pydantic tree that contains references.
Each backend supplies its own \texttt{OutputArchive}/\texttt{InputArchive} (which knows where to put bytes) and its own referencing type (which knows how to spell a pointer to those bytes).
Adding a fourth backend means writing those three pieces; no Python class needs to change.
Validation, schema export, and version stamping all live on the shared \texttt{ArchiveTree} layer, applied uniformly across backends.

\subsubsection{FITS Serialization}

\begin{figure}[ht]
  \begin{center}
    \begin{small}
      \begin{verbatim}
Filename: dp1.fits
No.    Name      Ver    Type      Cards   Dimensions   Format
  0  PRIMARY       1 PrimaryHDU     689   ()
  1  IMAGE         1 CompImageHDU     74   (4072, 4000)   float32
  2  MASK          1 CompImageHDU    109   (4072, 4000)   int32
  3  VARIANCE      1 CompImageHDU     74   (4072, 4000)   float32
  4  PSF/PIFF/INTERP/SOLUTION    1 BinTableHDU     12   1R x 1C   [9375D]
  5  JSON          1 BinTableHDU     11   1R x 1C   [PB(65720)]
  6  INDEX         1 BinTableHDU     23   6R x 7C   [24A, J, 8A, L, K, K, K]
      \end{verbatim}
    \end{small}
  \end{center}
  \caption{A DP1 visit image converted to the new FITS serialization.}
  \label{fig:new-fits}
\end{figure}

For FITS output we continue to write the image, mask, and variance to individual HDUs but the primary FITS header contains some additional keys to aid the deserialization tools.
These headers are:

\begin{description}
  \item[\textbf{INDXADDR}]
  The byte offset to the start of the HDU containing the HDU index.
  \item[\textbf{INDXSIZE}]
  The size of the index HDU in bytes.
  \item[\textbf{JSONADDR}]
  The byte offset to the start of the HDU containing the JSON data model.
  \item[\textbf{JSONSIZE}]
  The size of the JSON HDU.
\end{description}

These can be used to allow remote reading of a FITS file to allow the bare minimum of remote reads to determine the contents of the file.
We always try to write a FITS WCS to the headers of each image HDU, although this will sometimes have to be an approximation.

The \texttt{JSON} HDU contains the primary data model as JSON serialized from a Pydantic model.
When the model contains large arrays of numbers those are pulled out of the JSON serialization and stored in FITS binary table or image HDUs.
A reference to the FITS HDU is included in the JSON serialization so that on read the data can be located.

\subsubsection{HDF5 Serialization}

\begin{figure}[ht]
  \begin{center}
    \begin{small}
\begin{verbatim}
VisitImage  <NDF>

   DATA_ARRAY     <ARRAY>         {structure}
      DATA(4072,4000)  <_REAL>       -6.884751,-6.867677,-6.850569,-6.833529,-6.816523,
                                     ... -5.511999,-5.517168,-5.522304,-5.527539,-5.532708
      ORIGIN(2)      <_INT64>        0,0

   MORE           <EXT>           {structure}
      FITS(693)      <_CHAR*80>      'COMMENT   FITS (Flexible Image Transport System) format...'
                                     ... 'HIERARCH HAS-SIMULA...','EXPID   =          45508261376'
      LSST           <LSST>          {structure}
         JSON(1)        <_CHAR*88307>   '{"image":{"data":{"value":{"source":"ndf:/DATA_ARRAY/...'
         MASK           <NDF>           {structure}
            DATA_ARRAY     <ARRAY>         {structure}
               BAD_PIXEL      <_LOGICAL>      FALSE
               DATA(4072,4000,2)  <_UBYTE>    49,49,49,49,49,49,49,49,49,49,49,49,49,49,49,49,
                                              ... 0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0
               ORIGIN(3)      <_INT64>        0,0,0

            WCS            <WCS>           {structure}
               DATA(337)      <_CHAR*32>      ' Begin FrameSet',' Nframe = 3',' Currnt = 3',
                                              ' End UnitMap',' End CmpMap',' End FrameSet'

         PSF            <PSF>           {structure}
            PIFF           <PIFF>          {structure}
               INTERP         <INTERP>        {structure}
                  SOLUTION       <NDF>           {structure}
                     DATA_ARRAY     <ARRAY>         {structure}
                        DATA(15,625,1)  <_DOUBLE>      -8.80450720119882D-5,
                                                       ... 4.17704930562368D-14
                        ORIGIN(3)      <_INT64>        0,0,0

   QUALITY        <QUALITY>       {structure}
      BADBITS        <_UBYTE>        *
      QUALITY        <ARRAY>         {structure}
         BAD_PIXEL      <_LOGICAL>      FALSE
         DATA(4072,4000)  <_UBYTE>      1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,
                                        ... 1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1
         ORIGIN(2)      <_INTEGER>      0,0

   UNITS          <_CHAR*3>       'nJy'
   VARIANCE       <ARRAY>         {structure}
      DATA(4072,4000)  <_REAL>       272.9544,272.951,272.9477,272.9443,272.9409,272.9376,
                                     ... 259.7251,259.7245,259.7239,259.7233,259.7228,259.7221
      ORIGIN(2)      <_INT64>        0,0

   WCS            <WCS>           {structure}
      DATA(312)      <_CHAR*32>      ' Begin FrameSet',' Nframe = 3',' Base = 3',' Currnt = 2',
                                     ' Sft1 = 1',' Sft2 = 1',' End ShiftMap',' End FrameSet'

End of Trace.
\end{verbatim}
    \end{small}
  \end{center}
  \caption{The structure of the HDF5 representation of a DP1 visit FITS file converted to use the NDF data model. This is the output from the Starlink \texttt{hdstrace} command, which uses HDF5 group attributes to match the Starlink HDS semantics.}
  \label{fig:dp1-ndf}
\end{figure}

To demonstrate that the serialization interface was flexible and not FITS-specific, we decided to also write an HDF5 plugin.
We chose to use the Starlink N-Dimensional Data Format (NDF) data model\cite{2015A&C....12..146J} layered on the Starlink Hierarchical Data System (HDS) abstraction over HDF5.\cite{2015A&C....12..221J}

Our \texttt{MaskedImage} maps almost directly onto NDF: the image and variance arrays support pixel origin and units, and the WCS is represented by an AST native serialization.
The model also has a convention for including direct support for FITS header cards, stored as an extension in a character array.
There is one slight model mismatch concerning mask planes.
NDF has a ``quality'' array corresponding to the mask but is limited to only 8 mask planes, whereas we can support an arbitrary number.

For the more complex data models NDF provides sufficient flexibility to be able to represent them, although there is no native support for tabular data.
When there is additional information associated with a single image the convention is to place that in the extension ``airlock'' indicated by \texttt{.MORE}.
This section can contain arbitrary additional structure and is where we write the JSON model.
If there are multiple top-level images or masked images in the Python representation we would instead write those at the top level and the JSON component would also be at the top level since it is shared between all the images.

\subsubsection{VisitImage}

Fig.~\ref{fig:dp1-ndf} shows the NDF equivalent of the FITS file listed in Fig.~\ref{fig:new-fits}.
There is inherently more structure in the NDF representation.
Here a \texttt{QUALITY} component is created from a collapsed version of the full 32-bit mask, to enable standard Starlink applications to view a form of the mask.
The full mask is located at \texttt{MORE/LSST/MASK} and stores the 3-D version of the mask.
The Pydantic model itself is located at \texttt{MORE/LSST/JSON} and internally it references explicit numerical components of the main file.

\begin{lstlisting}[caption=PIFF solution in Pydantic model referencing a data component., language=json, label={lst:piff-ndf}]
"interp": {
  "tables": {
    "solution": {
      "metadata": {},
      "table": {
        "columns": [
          {
            "data": {
              "source": "ndf:/MORE/LSST/PSF/PIFF/INTERP/SOLUTION/DATA_ARRAY/DATA",
              "shape": [
                625,
                15
              ],
              "datatype": "float64",
              "byteorder": "big"
            },
            "name": "q"
          }
        ]
      }
    }
  }
}
\end{lstlisting}

Listing~\ref{lst:piff-ndf} shows the JSON serialization of part of the PIFF model and how it refers to the location of the corresponding data array in the HDF5 hierarchy.
The NDF writes a 3-D version of the array because there is an implicit counter from the row count.

\subsubsection{ColorImage}

\begin{figure}[ht]
  \begin{center}
    \begin{small}
\begin{verbatim}
ColorImage  <EXT>

   BLUE           <NDF>           {structure}
      DATA_ARRAY     <ARRAY>         {structure}
         DATA(8,5)      <_UBYTE>        180,195,233,192,127,216,137,62,43,114,26,78,157,128,
                                        ... 224,135,109,79,136,237,29,63,4,42,121,135,160,142,131
         ORIGIN(2)      <_INT64>        40,20

      WCS            <WCS>           {structure}
         DATA(187)      <_CHAR*32>      ' Begin FrameSet',' Nframe = 3',' Base = 3',' Currnt = 2',
                                        ... ' Sf...',' Sft2 = -19',' End ShiftMap',' End FrameSet'

   GREEN          <NDF>           {structure}
      DATA_ARRAY     <ARRAY>         {structure}
         DATA(8,5)      <_UBYTE>        209,3,200,106,54,88,31,99,143,45,112,77,157,156,36,
                                        ... 65,107,202,225,27,239,103,111,109,178,11,149,139,24
         ORIGIN(2)      <_INT64>        40,20

      WCS            <WCS>           {structure}
         DATA(187)      <_CHAR*32>      ' Begin FrameSet',' Nframe = 3',' Base = 3',' Currnt = 2',
                                        ... ' Sf...',' Sft2 = -19',' End ShiftMap',' End FrameSet'

   LSST           <LSST>           {structure}
      JSON(1)        <_CHAR*12703>   '{"red":{"data":{"source":"ndf:/RED/DATA_ARRAY/DATA","sha...'

   RED            <NDF>           {structure}
      DATA_ARRAY     <ARRAY>         {structure}
         DATA(8,5)      <_UBYTE>        86,172,68,196,13,110,180,176,133,126,47,187,116,228,
                                        ... 89,172,226,205,215,201,135,239,106,160,129,88,115,175
         ORIGIN(2)      <_INT64>        40,20

      WCS            <WCS>           {structure}
         DATA(187)      <_CHAR*32>      ' Begin FrameSet',' Nframe = 3',' Base = 3',' Currnt = 2',
                                        ... ' Sf...',' Sft2 = -19',' End ShiftMap',' End FrameSet'

End of Trace.
\end{verbatim}
    \end{small}
  \end{center}
  \caption{The structure of the HDF5 representation of a RGB color image serialized with the NDF data model.}
  \label{fig:color-ndf}
\end{figure}

A \texttt{ColorImage} is an image with WCS that represents three color channels.
In the NDF serialization, shown in Fig.~\ref{fig:color-ndf}, each color channel is an NDF stored at the top level and the data model is stored in the \texttt{/LSST/JSON} character array.
This demonstrates that we can use the hierarchy in multiple ways to group related concepts.

\section{Conclusion}

Our initial serialization system converting C++ data models to FITS that was developed in the early days of the LSST construction project\cite{2010SPIE.7740E..15A} did not meet our expectations in terms of cloud storage performance or ability to provide clear documentation to the community.
We have developed a new approach that uses bespoke classes for different image types to replace an approach where a single class has to represent all possible metadata extensions.
This has been coupled with an abstract file serialization system that lets us write the image data out in formats such as FITS and HDF5.
To aid with cloud performance we are also proposing a new FITS convention whereby we store an index containing the byte offsets of each HDU.
The intent is to switch to this new FITS format for Data Preview 2 which will be released in the second half of 2026.

\clearpage
